\chapter{Conclusions and outlook}

In this PhD project it was my goal to reveal structural and molecular details
about IL-10-GAG interaction with theoretical and computational means, and with
the help of experiments performed by collaborators. Furthermore, since the
\textit{in silico} methods for protein-GAG investigation are still in their
infancy, this thesis project included the development and improvement of
methodological approaches.

Application of a novel method for protein-GAG binding region analysis to IL-10
lead to the reliable prediction of a IL-10-GAG binding region centered on the
residues R102, R104, R106, and R107 of the human IL-10 monomer. Based on this
finding, the application of another novel method, Dynamic Molecular Docking
(DMD), in combination with a clustering method optimized for protein-GAG
systems, was able to derive further molecular detail about IL-10-GAG
interaction: it was found that R107 significantly stands out compared to all
other amino acid residues in IL-10 and supposedly plays a particularly important
role in IL-10-GAG recognition. Since this conclusion has been generated
\textit{in silico}, it is important to note that it is based on different
reliable types of data, including ensemble- and time-averaged quantities
obtained from molecular dynamics data collected on the microsecond time scale.
Additionally, it was revealed that K119 potentially plays an important role, as
it might cooperate with R107 in trapping a GAG from two sides simultaneously.
Application of DMD to the IL-10-GAG system led to the observation of two
principally different GAG binding orientations, which could both be valid in the
natural IL-10-GAG system.

Generally, it was found that GAGs do not bind to IL-10 with a binding affinity
as large as for instance in the FGF2-HP system. Still, especially with the help
of NMR data, it was revealed that IL-10-HP interaction occurs in a systematic
fashion, with a measurable impact on the HP backbone structure, as observed via
molecular dynamics-based HP structure modeling incorporating nuclear Overhauser
effect data obtained by our collaboration partners at the University Leipzig.

\nomenclature{SPR}{surface plasmon resonance}
\nomenclature{ITC}{isothermal titration calorimetry}

Behavioral differences among different GAG types were observed, in the sense
that all binding effects were more pronounced with an increasing number of
sulfate groups per disaccharide unit. This was shown especially in terms of the
hydrogen bonding data and the single-residue energy decomposition data as
obtained via DMD. However, the impact of \textit{certain} sulfate groups in the
IL-10-GAG system was so far not clarified using pure \textit{in silico}
investigation, so that corresponding observations made in NMR experiments and
also published by Salek-Ardakini et al. \cite{salek_ardakani_2000} could
unfortunately not be verified or structurally explained. While this type of
analysis is possible in the future using special \textit{in silico} sampling
techniques such as DMD, an absolutely certain knowledge about the GAG binding
site should be available from experiments --- for the sake of creating
\textit{reliable} results. Hence, more experimental data is required with
respect to the IL-10-GAG system for the success of further \textit{in silico}
investigations. After what has been found out so far, the obvious experimental
tasks are now the investigation of the role of R107 for GAG-binding affinity in
NMR, SPR or ITC experiments, and the role of R107 for the impact of GAG-binding
on IL-10 biology via e.g.\ \textit{in vitro} cell proliferation experiments,
whereas all of these studies do require the creation of corresponding IL-10
mutants.

The kind of data obtained about the IL-10-GAG system so far allows two realistic
scenarios in which GAG-binding to IL-10 affects its biological function. In one
scenario, long GAG chains impair the diffusion of the IL-10 cytokine and
therefore may take part in a complex cytokine concentration regulation system.
In the second scenario, GAG-binding to IL-10 sterically interferes with the
binding of IL-10R2 to IL-10, which could directly inhibit the activation of the
IL-10-triggered cell-internal signaling cascade. Clearly, more reliable data
about the structure of the ternary IL-10 signaling complex than current models
provide are required to obtain further insights in this regard. Currently, the
cooperative binding behavior in which one GAG molecule interacts with two
separate binding regions simultaneously (observed via NMR experiments
\cite{kuenze_gehrcke_2014}) fits best the idea of polymeric GAGs catching IL-10
in the groove of the V-shape and impairing its diffusion.

While the insights into the molecular mechanism of the IL-10-GAG system obtained
in the course of this project might be slightly disappointing, the yield in
novel \textit{in silico} methods for the generic investigation of protein-GAG
systems can be considered rather successful. The binding region prediction
method presented in \cref{chapter:bspred} provides a valuable bare-bones tool
for future investigators of protein-GAG systems, with clear advantages over
existing methods. The molecular structure clustering method described in
\cref{chapter:clustering} is based on a solid concept, involves components
heavily optimized for protein-GAG systems, and serves as a useful and sometimes
\textit{essential} data post-processing technique. Dynamic Molecular Docking
(DMD) can be considered the most valuable core result of this thesis project: it
is an advanced \textit{in silico} method for the investigation of protein-GAG
systems, with one clear goal: yielding reliable data based on large-scale
molecular dynamics simulations. DMD could largely facilitate and improve the
investigation of protein-GAG systems in the future, and renders a perfect triad
together with the newly developed protein-GAG binding region prediction and
docking solution clustering.
